Analysis of the time-dependent response to Estradiol through ChIP-Seq data

### 2018-11-26 S2_ChIP-seq_ER/rmarkdown.Rmd
updated ferderico's code

---
title: "Analysis of the time-dependent response to Estradiol through ChIP-Seq data and the VULCAN package"
author: "Federico M. Giorgi"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    fig_caption: yes
    fig_height: 8
    fig_width: 8
    toc: true # turns on table of contents
    toc_depth: 3  # up to three depths of headings (specified by #, ## and ###)
    number_sections: true  # if you want number sections at each table header
    theme: united
    highlight: tango
---

```{r setup_vulcan, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(fig.align="center")
knitr::opts_chunk$set(fig.path = "rmarkdown_images/")
par.backup<-par()
```

```{r captions, include=FALSE}
set.seed(1)
#Figure and Table Caption Numbering, for HTML do it manually
captabNo <- 1; capfigNo <- 1;

@@

#Function to add the Figure Number
capfig <- function(x){
  #x <- paste0("Figure S",capfigNo,". ",x)
  #capfigNo <<- capfigNo + 1
  x
}
```

## Introduction
Vulcan (VirtUaL ChIP-Seq Analysis through Networks) is a pipeline that combines
ChIP-Seq data and regulatory networks to obtain transcription factors that are
likely affected by a specific stimulus. In order to do so, our package combines
strategies from different BioConductor packages: \emph{DESeq} for data
normalization, \emph{ChIPpeakAnno} and \emph{DiffBind} for annotation and
definition of ChIP-Seq genomic peaks, \emph{csaw} to define optimal peak width
and \emph{viper} for applying a regulatory network over a differential binding
signature.
Usage of gene regulatory networks to analyze biological systems has witnessed
an exponential increase in the last decade, due to the ease of obtaining
genome-wide expression data (Margolin et al., 2005; Giorgi et al., 2013; Castro
et al., 2015). Recently, the VIPER approach to interrogate these network models
has been proposed to infer transcription factor activity using the expression
of a collection of their putative targets, i.e. their regulon (Alvarez et al.,
2016). In the VIPER algorithm, gene-level differential expression signatures
are obtained for either individual samples (relative to the mean of the dataset)
or between groups of samples, and regulons are tested for enrichment. Ideally,
TF regulons can be tested on promoter-level differential binding signatures
generated from ChIP-Seq experiments, in order to ascertain the global change
in promoter occupancy for gene sets. In our study, we propose an extension
of the VIPER algorithm to specifically analyze TF occupancy in ChIP-Seq
experiments. Our VULCAN algorithm uses ChIP-Seq data obtained for a given TF
to provide candidate coregulators of the response to a given stimulus.
The analysis is based on identifying differentially bound genes and testing
their enrichment in the regulon of potential co-regulatory factors.

## Software setup
VULCAN is conveniently provided as an R package available from the Bioconductor repository.
```{r, message=FALSE, warning=FALSE}
#source("http://bioconductor.org/biocLite.R")
#biocLite("vulcan")
library(vulcan)
```
Other packages are required for the execution of the analysis and the visualization of the results. A \textit{results} folder will be also created to store intermediate steps of the analysis.
```{r, message=FALSE, warning=FALSE}
library(gplots) # for heatmap.2
library(org.Hs.eg.db)
library(gridExtra) # for the pathway part
library(GeneNet) # to generate partial correlation networks
if(!file.exists("results")){dir.create("results")}
```
Finally, we will define here some convenience functions, such as those to convert from gene symbols to entrez ids, and viceversa.
```{r, message=FALSE, warning=FALSE}
list_eg2symbol<-as.list(org.Hs.egSYMBOL[mappedkeys(org.Hs.egSYMBOL)])
e2s<-function(ids){
  ids <- as.character(ids)
  outlist <- list_eg2symbol[ids]
  names(outlist) <- ids
  outlist[is.na(outlist)] <- paste("unknown.", ids[is.na(outlist)], sep = "")
  outlist <- gsub("unknown.unknown.", "", outlist)
  return(outlist)
}
list_symbol2eg <- as.character(org.Hs.egSYMBOL2EG[mappedkeys(org.Hs.egSYMBOL2EG)])
s2e<-function(ids){
  ids <- as.character(ids)
  outlist <- list_symbol2eg[ids]
  names(outlist) <- ids
  outlist[is.na(outlist)] <- paste("unknown.", ids[is.na(outlist)], sep = "")
  outlist <- gsub("unknown.unknown.", "", outlist)
  return(outlist)
}
```

## Import ChIP-Seq data using Vulcan
ChIP-Seq data was generated using a cell line model of ER+ breast cancer, MCF7, at 0’, 45’ and 90’ after stimulation with estradiol. The binding profile of ER at each timepoint was then compared between time points using differential binding analysis. From the temporal comparison ER binding we established four classes of binding pattern: early responders, repressed transcription factors, late responders and candidate cyclic genes.

@@

### Initial import
VULCAN requires an input sheet file in CSV format describing the samples and the location of the individual input files
```{r, message=FALSE, warning=FALSE}
sheetfile<-"chipseq/holding/sheet.csv"

@@

### Annotation
VULCAN identifies the location of each peak in each sample, and according to the selected method, assigns a score to each gene promoter. There are a few methods available.

@@

### Normalization
At this step, genes are quantified according to the number of reads that could be associated to their promoters. The algorithms within VULCAN (\textit{viper} and \textit{DESeq2}) require however data to be normalized via Variance-Stabilizing Transformation (Anders and Huber, 2010).

@@

### Dataset Sample Clustering
Here, we show how the samples cluster together using peak raw counts, VST-normalized peak raw counts and peak RPKMs. In the following heatmaps, there are `r nrow(vobj$rawcounts)` rows, one per gene promoter, and 16 columns, one per sample. The sample name indicates the time point in minutes (T0, T45 or T90) and the number of replicate (_1, _2, _3, _4).
```{r, fig.cap=capfig('Dataset clustering with Raw Counts'), message=FALSE, warning=FALSE}
heatmap.2(vobj$rawcounts,scale="row",
          col=colorpanel(1000,"navy","white","red3"),tracecol="black",labRow="")
mtext("Raw Counts")
```

@@

### MA plots
Without changing the format of the input data, we can use the Bioconductor DiffBind package to visualize the amplitude of changes in ER binding between time points. One way to do this is an MA plot, which shows the differences between measurements taken in two groups (e.g. 45' vs 00'), by transforming the promoter peak intensity data onto M (log ratio) and A (mean average) scales.

@@

## Apply a coregulatory network over a ChIP-Seq profile
Once the data has been loaded, VULCAN applies a regulatory network over differential binding signatures to define Transcription Factors whose networks are most affected by the treatment. In our analysis, we will be using three different networks generated via ARACNe (Margolin et al., 2006): two are breast cancer-specific and are derived from the TCGA and METABRIC data collection, respectively. A third network is used as a negative control, highlighting regulatory mechanisms derived from the Amyloid Leukemia dataset (TCGA).
### Loading the input networks and processed data
First, we will load the output from the previous paragraphs (chipseq data annotated and normalized) and the three networks.
```{r, message=FALSE, warning=FALSE}
# Load imported vulcan object
load("results/001_vobj.rda")

@@

load("networks/metabric-regulon-tfs.rda")
metabric_regulon<-regulon
rm(regulon)
```
Accessing the vulcan object /texit{vobj} can give us informations on the sample groups thereby contained.
```{r, warning=FALSE}
names(vobj$samples)
```
### Running the core Vulcan function
The final Vulcan pipeline step requires three input objects:

@@

```{r, message=FALSE, warning=FALSE}
fname<-"results/002_vobj_networks.rda"
list_eg2symbol<-as.list(org.Hs.egSYMBOL[mappedkeys(org.Hs.egSYMBOL)]) 
if(!file.exists(fname)){
  vobj_tcga_90<-vulcan(vobj,network=tcga_regulon,contrast=c("t90","t0"),annotation=list_eg2symbol)
  vobj_tcga_45<-vulcan(vobj,network=tcga_regulon,contrast=c("t45","t0"),annotation=list_eg2symbol)
  vobj_metabric_90<-vulcan(vobj,network=metabric_regulon,contrast=c("t90","t0"),annotation=list_eg2symbol)
  vobj_metabric_45<-vulcan(vobj,network=metabric_regulon,contrast=c("t45","t0"),annotation=list_eg2symbol)
  vobj_negative_90<-vulcan(vobj,network=laml_regulon,contrast=c("t90","t0"),annotation=list_eg2symbol)
  vobj_negative_45<-vulcan(vobj,network=laml_regulon,contrast=c("t45","t0"),annotation=list_eg2symbol)
  save(
    vobj_tcga_90,
    vobj_tcga_45,
    vobj_metabric_90,
    vobj_metabric_45,
    vobj_negative_90,
    vobj_negative_45,
    file=fname
  )
} else {
  load(fname)
}
```
### Visualize the relative TF activity
Vulcan has now generated activity scores for each Transcription Factor, which specify the global ER binding strength on the promoters of their targets.

#### METABRIC results

@@

text(3,metabricmat["ESR1",3],label="ESR1",pos=4,cex=1,font=2,col="red3")

@@

These TFs can be grouped in classes. For example, TFs whose activity is already significant after 45mins and remains significant at 90 minutes after estradiol treatment can be dubbed **early responders**.

@@

matplot(t(metabricmat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Early responders",xlim=c(1,3.3),ylim=c(0,max(metabricmat)))

@@

newcoords<-setNames(seq(min(oricoords)*0.7,max(oricoords),length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.8,font=2)
text(3,newcoords["ESR1"],label="ESR1",pos=4,cex=1,font=2,col="red3")

@@

TFs whose *repressed* targets are bound will yield a negative activity score. These **repressed TFs** are symmetrically opposite to early responder TFs.

@@

matplot(t(metabricmat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Repressed TFs",xlim=c(1,3.3),ylim=c(min(metabricmat),0))

@@

newcoords<-setNames(seq(min(oricoords),max(oricoords)*0.5,length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.8,font=2)

@@

Some TFs appear to have their targets bound at 45 minutes, but then unoccupied at 90 minutes. This "updown" behavior is consistent to what observed in previous literature about the cyclic properties of certain components of the ER DNA-binding complex, and therefore we dubbed them **candidate cyclic TFs**.

@@

matplot(t(metabricmat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Candidate Cyclic TFs",xlim=c(1,3.3),ylim=c(min(metabricmat),max(metabricmat)))

@@

text(3,newcoords,label=names(oricoords),pos=4,cex=0.8,font=2)

@@

Finally, a category of TFs appear to be activated but at 90 minutes only. We call this category of TFs **late responders**.

@@

matplot(t(metabricmat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Late responders",xlim=c(1,3.3),ylim=c(0,max(metabricmat)))

@@

newcoords<-setNames(seq(min(oricoords)*0.7,min(oricoords)*1.4,length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.8,font=2)

@@

#### TCGA results
All the TFs and the four categories of TFs were inferred also using the TCGA network.

@@

```{r, fig.cap=capfig("Global TF activity after Estradiol Treatment in MCF7 cells, inferred using the TCGA network, highlighting TFs significantly upregulated at 45 minutes and 90 minutes"), message=FALSE, warning=FALSE,dev.args=list(pointsize=14)}
threshold<-p2z(0.05)
tfclass<-tfs[tcgamat[,"T45"]>=threshold&tcgamat[,"T90"]>=threshold]
matplot(t(tcgamat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Early responders",xlim=c(1,3.3),ylim=c(0,max(tcgamat)))
axis(1,at=c(1:3),labels=colnames(tcgamat))
abline(h=c(0,threshold,-threshold),lty=2)
matplot(t(tcgamat[tfclass,]),type="l",col="red3",lty=1,lwd=2,add=TRUE)
# Repel a bit
plabels<-tfclass
oricoords<-sort(setNames(tcgamat[plabels,3],plabels))
newcoords<-setNames(seq(min(oricoords),max(oricoords),length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)
text(3,newcoords["ESR1"],label="ESR1",pos=4,cex=0.6,font=2,col="red3")
mtext("TCGA network")
```

```{r, fig.cap=capfig("Global TF activity after Estradiol Treatment in MCF7 cells, inferred using the TCGA network, highlighting TFs significantly downregulated at 45 minutes and 90 minutes"), message=FALSE, warning=FALSE,dev.args=list(pointsize=14)}
threshold<-p2z(0.05)
tfclass<-tfs[tcgamat[,"T45"]<=-threshold&tcgamat[,"T90"]<=-threshold]
matplot(t(tcgamat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Repressed TFs",xlim=c(1,3.3),ylim=c(min(tcgamat),0))
axis(1,at=c(1:3),labels=colnames(tcgamat))
abline(h=c(0,threshold,-threshold),lty=2)
matplot(t(tcgamat[tfclass,]),type="l",col="cornflowerblue",lty=1,lwd=2,add=TRUE)
# Repel a bit
plabels<-tfclass
oricoords<-sort(setNames(tcgamat[plabels,3],plabels))
newcoords<-setNames(seq(min(oricoords),max(oricoords),length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)
mtext("TCGA network")
```

```{r, fig.cap=capfig("Global TF activity after Estradiol Treatment in MCF7 cells, inferred using the TCGA network, highlighting TFs significantly upregulated at 45 minutes but not at 90 minutes"), message=FALSE, warning=FALSE,dev.args=list(pointsize=14)}
threshold<-p2z(0.05)
tfclass<-tfs[tcgamat[,"T45"]>=threshold&tcgamat[,"T90"]<=threshold]
matplot(t(tcgamat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Candidate Cyclic TFs",xlim=c(1,3.3),ylim=c(0,max(tcgamat)))
axis(1,at=c(1:3),labels=colnames(tcgamat))
abline(h=c(0,threshold,-threshold),lty=2)
matplot(t(tcgamat[tfclass,]),type="l",col="seagreen",lty=1,lwd=2,add=TRUE)
# Repel a bit
plabels<-tfclass
oricoords<-sort(setNames(tcgamat[plabels,3],plabels))
newcoords<-setNames(seq(min(oricoords),max(oricoords),length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)
mtext("TCGA network")
```

```{r, fig.cap=capfig("Global TF activity after Estradiol Treatment in MCF7 cells, inferred using the TCGA network, highlighting TFs significantly upregulated at 90 minutes but not at 45 minutes"), message=FALSE, warning=FALSE,dev.args=list(pointsize=14)}
## Dn-Up TF class
threshold<-p2z(0.05)
tfclass<-tfs[tcgamat[,"T45"]<=threshold&tcgamat[,"T45"]>0&tcgamat[,"T90"]>=threshold]
matplot(t(tcgamat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Late responders",xlim=c(1,3.3),ylim=c(0,max(tcgamat)))
axis(1,at=c(1:3),labels=colnames(tcgamat))
abline(h=c(0,threshold,-threshold),lty=2)
matplot(t(tcgamat[tfclass,]),type="l",col="darkorange",lty=1,lwd=2,add=TRUE)
# Repel a bit
plabels<-tfclass
oricoords<-sort(setNames(tcgamat[plabels,3],plabels))
newcoords<-setNames(seq(min(oricoords),min(oricoords)*1.2,length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)
mtext("TCGA network")
```

>>>>

---
title: Analysis of the time-dependent response to Estradiol through ChIP-Seq data
  and the VULCAN package
author: "Federico M. Giorgi"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    fig_height: 8
    fig_width: 8
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 3
fig_caption: yes
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(fig.align="center")
knitr::opts_chunk$set(fig.path = "rmarkdown_images/")
par.backup<-par()
```

```{r captions, include=FALSE, message=FALSE}
set.seed(1)
#Figure and Table Caption Numbering, for HTML do it manually
captabNo <- 1; capfigNo <- 1;

@@

#Function to add the Figure Number
capfig <- function(x){
  x <- paste0("Figure S",capfigNo,". ",x)
  capfigNo <<- capfigNo + 1
  x
}
```

# Introduction
Vulcan (VirtUaL ChIP-Seq Analysis through Networks) is a pipeline that combines
ChIP-Seq data and regulatory networks to obtain transcription factors that are
likely affected by a specific stimulus. In order to do so, our package combines
strategies from different BioConductor packages: \emph{DESeq} for data
normalization, \emph{ChIPpeakAnno} and \emph{DiffBind} for annotation and
definition of ChIP-Seq genomic peaks, \emph{csaw} to define optimal peak width
and \emph{viper} for applying a regulatory network over a differential binding
signature.
Usage of gene regulatory networks to analyze biological systems has witnessed
an exponential increase in the last decade, due to the ease of obtaining
genome-wide expression data (Margolin et al., 2005; Giorgi et al., 2013; Castro
et al., 2015). Recently, the VIPER approach to interrogate these network models
has been proposed to infer transcription factor activity using the expression
of a collection of their putative targets, i.e. their regulon (Alvarez et al.,
2016). In the VIPER algorithm, gene-level differential expression signatures
are obtained for either individual samples (relative to the mean of the dataset)
or between groups of samples, and regulons are tested for enrichment. Ideally,
TF regulons can be tested on promoter-level differential binding signatures
generated from ChIP-Seq experiments, in order to ascertain the global change
in promoter occupancy for gene sets. In our study, we propose an extension
of the VIPER algorithm to specifically analyze TF occupancy in ChIP-Seq
experiments. Our VULCAN algorithm uses ChIP-Seq data obtained for a given TF
to provide candidate coregulators of the response to a given stimulus.
The analysis is based on identifying differentially bound genes and testing
their enrichment in the regulon of potential co-regulatory factors.

# Software setup
VULCAN is conveniently provided as an R package available from the Bioconductor repository.
```{r, message=FALSE, warning=FALSE}
#source("http://bioconductor.org/biocLite.R")
#biocLite("vulcan")
library(vulcan)
```
Other packages are required for the execution of the analysis and the visualization of the results. A \textit{results} folder will be also created to store intermediate steps of the analysis.
```{r, message=FALSE, warning=FALSE}
library(gplots) # for heatmap.2
library(org.Hs.eg.db)
library(gridExtra) # for the pathway part
library(GeneNet) # to generate partial correlation networks
if(!file.exists("results")){dir.create("results")}
```
Finally, we will define here some convenience functions, such as those to convert from gene symbols to entrez ids, and viceversa.
```{r, message=FALSE, warning=FALSE}
list_eg2symbol<-as.list(org.Hs.egSYMBOL[mappedkeys(org.Hs.egSYMBOL)])
e2s<-function(ids){
  ids <- as.character(ids)
  outlist <- list_eg2symbol[ids]
  names(outlist) <- ids
  outlist[is.na(outlist)] <- paste("unknown.", ids[is.na(outlist)], sep = "")
  outlist <- gsub("unknown.unknown.", "", outlist)
  return(outlist)
}
list_symbol2eg <- as.character(org.Hs.egSYMBOL2EG[mappedkeys(org.Hs.egSYMBOL2EG)])
s2e<-function(ids){
  ids <- as.character(ids)
  outlist <- list_symbol2eg[ids]
  names(outlist) <- ids
  outlist[is.na(outlist)] <- paste("unknown.", ids[is.na(outlist)], sep = "")
  outlist <- gsub("unknown.unknown.", "", outlist)
  return(outlist)
}
```

# Import ChIP-Seq data using Vulcan
ChIP-Seq data was generated using a cell line model of ER+ breast cancer, MCF7, at 0’, 45’ and 90’ after stimulation with estradiol. The binding profile of ER at each timepoint was then compared between time points using differential binding analysis. From the temporal comparison ER binding we established four classes of binding pattern: early coactivators, early corepressors, late coactivators and transient coactivators.

@@

## Initial import
VULCAN requires an input sheet file in CSV format describing the samples and the location of the individual input files
```{r, message=FALSE, warning=FALSE}
sheetfile<-"chipseq/holding/sheet.csv"

@@

## Annotation
VULCAN identifies the location of each peak in each sample, and according to the selected method, assigns a score to each gene promoter. There are a few methods available.

@@

## Normalization
At this step, genes are quantified according to the number of reads that could be associated to their promoters. The algorithms within VULCAN (\textit{viper} and \textit{DESeq2}) require however data to be normalized via Variance-Stabilizing Transformation (Anders and Huber, 2010).

@@

## Dataset Sample Clustering
Here, we show how the samples cluster together using peak raw counts, VST-normalized peak raw counts and peak RPKMs. In the following heatmaps, there are `r nrow(vobj$rawcounts)` rows, one per gene promoter, and 16 columns, one per sample. The sample name indicates the time point in minutes (T0, T45 or T90) and the number of replicate (_1, _2, _3, _4).
```{r, fig.cap=capfig('Dataset clustering with Raw Counts'), message=FALSE, warning=FALSE}
heatmap.2(vobj$rawcounts,scale="row",
          col=colorpanel(1000,"navy","white","red3"),tracecol="black",labRow="")
mtext("Raw Counts")
```

@@

## MA plots
Without changing the format of the input data, we can use the Bioconductor DiffBind package to visualize the amplitude of changes in ER binding between time points. One way to do this is an MA plot, which shows the differences between measurements taken in two groups (e.g. 45' vs 00'), by transforming the promoter peak intensity data onto M (log ratio) and A (mean average) scales.

@@

# Apply a coregulatory network over a ChIP-Seq profile
Once the data has been loaded, VULCAN applies a regulatory network over differential binding signatures to define Transcription Factors whose networks are most affected by the treatment. In our analysis, we will be using three different networks generated via ARACNe (Margolin et al., 2006): two are breast cancer-specific and are derived from the TCGA and METABRIC data collection, respectively. A third network is used as a negative control, highlighting regulatory mechanisms derived from the Amyloid Leukemia dataset (TCGA).
## Loading the input networks and processed data
First, we will load the output from the previous paragraphs (chipseq data annotated and normalized) and the three networks.
```{r, message=FALSE, warning=FALSE}
# Load imported vulcan object
load("results/001_vobj.rda")

@@

load("networks/metabric-regulon-tfs.rda")
metabric_regulon<-regulon
rm(regulon)
```
Accessing the vulcan object /texit{vobj} can give us informations on the sample groups thereby contained.
```{r, warning=FALSE}
names(vobj$samples)
```
## Running the core Vulcan function
The final Vulcan pipeline step requires three input objects:

@@

```{r, message=FALSE, warning=FALSE}
fname<-"results/002_vobj_networks.rda"
list_eg2symbol<-as.list(org.Hs.egSYMBOL[mappedkeys(org.Hs.egSYMBOL)]) 
if(!file.exists(fname)){
  vobj_tcga_90<-vulcan(vobj,network=tcga_regulon,contrast=c("t90","t0"),annotation=list_eg2symbol)
  vobj_tcga_45<-vulcan(vobj,network=tcga_regulon,contrast=c("t45","t0"),annotation=list_eg2symbol)
  vobj_metabric_90<-vulcan(vobj,network=metabric_regulon,contrast=c("t90","t0"),annotation=list_eg2symbol)
  vobj_metabric_45<-vulcan(vobj,network=metabric_regulon,contrast=c("t45","t0"),annotation=list_eg2symbol)
  vobj_negative_90<-vulcan(vobj,network=laml_regulon,contrast=c("t90","t0"),annotation=list_eg2symbol)
  vobj_negative_45<-vulcan(vobj,network=laml_regulon,contrast=c("t45","t0"),annotation=list_eg2symbol)
  save(
    vobj_tcga_90,
    vobj_tcga_45,
    vobj_metabric_90,
    vobj_metabric_45,
    vobj_negative_90,
    vobj_negative_45,
    file=fname
  )
} else {
  load(fname)
}
```
## Visualize the relative TF activity
Vulcan has now generated activity scores for each Transcription Factor, which specify the global ER binding strength on the promoters of their targets.

### METABRIC results

@@

text(3,metabricmat["ESR1",3],label="ESR1",pos=4,cex=0.6,font=2,col="red3")

@@

These TFs can be grouped in classes. For example, TFs whose activity is already significant after 45mins and remains significant at 90 minutes after estradiol treatment can be dubbed **early coactivators**.

@@

matplot(t(metabricmat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Early coactivators",xlim=c(1,3.3),ylim=c(0,max(metabricmat)))

@@

text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)
text(3,newcoords["ESR1"],label="ESR1",pos=4,cex=0.6,font=2,col="red3")

@@

TFs whose *repressed* targets are bound will yield a negative activity score. These **early corepressors** are symmetrically opposite to early responder TFs.

@@

matplot(t(metabricmat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Early corepressors",xlim=c(1,3.3),ylim=c(min(metabricmat),0))

@@

text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)

@@

Some TFs appear to have their targets bound at 45 minutes, but then unoccupied at 90 minutes. This "updown" behavior is consistent to what observed in previous literature about the cyclic properties of certain components of the ER DNA-binding complex, and therefore we dubbed them **transient coactivators**.

@@

matplot(t(metabricmat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Transient coactivators",xlim=c(1,3.3),ylim=c(min(metabricmat),max(metabricmat)))

@@

text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)

@@

Finally, a category of TFs appear to be activated but at 90 minutes only. We call this category of TFs **late coactivators**.

@@

matplot(t(metabricmat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Late coactivators",xlim=c(1,3.3),ylim=c(0,max(metabricmat)))

@@

newcoords<-setNames(seq(min(oricoords),min(oricoords)*1.2,length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)

@@

### TCGA results
All the TFs and the four categories of TFs were inferred also using the TCGA network.

@@

```{r, fig.cap=capfig("Global TF activity after Estradiol Treatment in MCF7 cells, inferred using the TCGA network, highlighting TFs significantly upregulated at 45 minutes and 90 minutes"), message=FALSE, warning=FALSE,dev.args=list(pointsize=14)}
threshold<-p2z(0.05)
tfclass<-tfs[tcgamat[,"T45"]>=threshold&tcgamat[,"T90"]>=threshold]
matplot(t(tcgamat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Early coactivators",xlim=c(1,3.3),ylim=c(0,max(tcgamat)))
axis(1,at=c(1:3),labels=colnames(tcgamat))
abline(h=c(0,threshold,-threshold),lty=2)
matplot(t(tcgamat[tfclass,]),type="l",col="red3",lty=1,lwd=2,add=TRUE)
# Repel a bit
plabels<-tfclass
oricoords<-sort(setNames(tcgamat[plabels,3],plabels))
newcoords<-setNames(seq(min(oricoords),max(oricoords),length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)
text(3,newcoords["ESR1"],label="ESR1",pos=4,cex=0.6,font=2,col="red3")
mtext("TCGA network")
```

```{r, fig.cap=capfig("Global TF activity after Estradiol Treatment in MCF7 cells, inferred using the TCGA network, highlighting TFs significantly downregulated at 45 minutes and 90 minutes"), message=FALSE, warning=FALSE,dev.args=list(pointsize=14)}
threshold<-p2z(0.05)
tfclass<-tfs[tcgamat[,"T45"]<=-threshold&tcgamat[,"T90"]<=-threshold]
matplot(t(tcgamat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Early corepressors",xlim=c(1,3.3),ylim=c(min(tcgamat),0))
axis(1,at=c(1:3),labels=colnames(tcgamat))
abline(h=c(0,threshold,-threshold),lty=2)
matplot(t(tcgamat[tfclass,]),type="l",col="cornflowerblue",lty=1,lwd=2,add=TRUE)
# Repel a bit
plabels<-tfclass
oricoords<-sort(setNames(tcgamat[plabels,3],plabels))
newcoords<-setNames(seq(min(oricoords),max(oricoords),length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)
mtext("TCGA network")
```

```{r, fig.cap=capfig("Global TF activity after Estradiol Treatment in MCF7 cells, inferred using the TCGA network, highlighting TFs significantly upregulated at 45 minutes but not at 90 minutes"), message=FALSE, warning=FALSE,dev.args=list(pointsize=14)}
threshold<-p2z(0.05)
tfclass<-tfs[tcgamat[,"T45"]>=threshold&tcgamat[,"T90"]<=threshold]
matplot(t(tcgamat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Transient coactivators",xlim=c(1,3.3),ylim=c(0,max(tcgamat)))
axis(1,at=c(1:3),labels=colnames(tcgamat))
abline(h=c(0,threshold,-threshold),lty=2)
matplot(t(tcgamat[tfclass,]),type="l",col="seagreen",lty=1,lwd=2,add=TRUE)
# Repel a bit
plabels<-tfclass
oricoords<-sort(setNames(tcgamat[plabels,3],plabels))
newcoords<-setNames(seq(min(oricoords),max(oricoords),length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)
mtext("TCGA network")
```

```{r, fig.cap=capfig("Global TF activity after Estradiol Treatment in MCF7 cells, inferred using the TCGA network, highlighting TFs significantly upregulated at 90 minutes but not at 45 minutes"), message=FALSE, warning=FALSE,dev.args=list(pointsize=14)}
## Dn-Up TF class
threshold<-p2z(0.05)
tfclass<-tfs[tcgamat[,"T45"]<=threshold&tcgamat[,"T45"]>0&tcgamat[,"T90"]>=threshold]
matplot(t(tcgamat),type="l",col="grey",ylab="VULCAN NES",xaxt="n",lty=1,main="Late coactivators",xlim=c(1,3.3),ylim=c(0,max(tcgamat)))
axis(1,at=c(1:3),labels=colnames(tcgamat))
abline(h=c(0,threshold,-threshold),lty=2)
matplot(t(tcgamat[tfclass,]),type="l",col="darkorange",lty=1,lwd=2,add=TRUE)
# Repel a bit
plabels<-tfclass
oricoords<-sort(setNames(tcgamat[plabels,3],plabels))
newcoords<-setNames(seq(min(oricoords),min(oricoords)*1.2,length.out=length(oricoords)),names(oricoords))
text(3,newcoords,label=names(oricoords),pos=4,cex=0.6,font=2)
mtext("TCGA network")
```

### Comparing TCGA and METABRIC results
The following scatterplots compare the activity inferred by VULCAN using two alternative networks, derived from the TCGA and METABRIC breast cancer datasets using ARACNe-AP (Giorgi et al., 2016) with default parameters.
```{r, fig.cap=capfig("Comparing TF activities inferred by VULCAN using two different breast cancer dataset-derived networks on the ER 45 minutes signature. PCC indicates the Pearson Correlation Coefficient"), message=FALSE, warning=FALSE,dev.args=list(pointsize=16)}
common<-intersect(rownames(tcgamat),rownames(metabricmat))
set.seed(1)
x<-tcgamat[common,"T45"]
y<-metabricmat[common,"T45"]
plot(x,y,xlab="TCGA, 45mins",ylab="METABRIC, 45mins",pch=20,col="grey",xlim=c(min(x)*1.2,max(x)*1.2))
grid()
#toshow<-c("ESR1","GATA3","RARA","HSF1")
toshow<-names(which(rank(rank(-abs(x))+rank(-abs(y)))<=20))
#toshow<-unique(c(toshow,names(sort(rank(-abs(x))[1:10])),names(sort(rank(-abs(y)))[1:10])))
textplot2(x[toshow],y[toshow],words=toshow,new=FALSE,cex=0.8)
lm1<-lm(y~x)
abline(lm1$coef)
pcc<-cor.test(x,y)
mtext(paste0("PCC=",signif(pcc$estimate,2)," (p=",signif(pcc$p.value,2),")"))
```
```{r, fig.cap=capfig("Comparing TF activities inferred by VULCAN using two different breast cancer dataset-derived networks on the ER 90 minutes signature. PCC indicates the Pearson Correlation Coefficient"), message=FALSE, warning=FALSE,dev.args=list(pointsize=16)}
x<-tcgamat[common,"T90"]
y<-metabricmat[common,"T90"]
plot(x,y,xlab="TCGA, 90mins",ylab="METABRIC, 90mins",pch=20,col="grey",xlim=c(min(x)*1.2,max(x)*1.2))
grid()
#toshow<-c("ESR1","GATA3","RARA","HSF1")
toshow<-names(which(rank(rank(-abs(x))+rank(-abs(y)))<=20))
#toshow<-unique(c(toshow,names(sort(rank(-abs(x))[1:10])),names(sort(rank(-abs(y)))[1:10])))
textplot2(x[toshow],y[toshow],words=toshow,new=FALSE,cex=0.8)
lm1<-lm(y~x)
abline(lm1$coef)
pcc<-cor.test(x,y)
mtext(paste0("PCC=",signif(pcc$estimate,2)," (p=",signif(pcc$p.value,2),")"))
abline(lm1$coef)
```

### Testing with a different context network as a negative control
Regulatory networks can be tissue-specific, and so we built a third network using the same parameters as the two breast cancer data-based ones. The third network is built on leukemic samples from the TCGA AML dataset. Our results show a weaker activity for all TFs, which is only weakly correlated to that inferred via the TCGA breast cancer network.
```{r, fig.cap=capfig("Comparison between activities inferred through a breast cancer TCGA dataset and the AML dataset. PCC indicates the Pearson Correlation Coefficient"), message=FALSE, warning=FALSE}
negative_90=vobj_negative_90$mrs[,"NES"]
negative_45=vobj_negative_45$mrs[,"NES"]
tfs<-names(negative_45)
negativemat<-cbind(rep(0,length(negative_45)),negative_45[tfs],negative_90[tfs])
colnames(negativemat)<-c("T0","T45","T90")
